Unshuffling Permutations

### 2015-09-25 submission.tex
typos
N1, N2, N3, N4
introduction
Typos

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product}, denoted $\shuffle$, is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word.
Besides purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
\begin{equation}
    O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
\begin{equation}
    O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
\end{equation}
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

@@

We follow the usual terminology on words~\cite{ChoffrutKarhumaki1997}.
Let us recall here the most important ones. Let $u$ be a word. The
length of $u$ is denoted by $|u|$. The {\em empty word}, the only word
of null length, is denoted by $\epsilon$. We denote by $\widetilde{u}$
the {\em mirror image} of $u$, that is the word
$u_{|u|} u_{|u| - 1} \dots u_1$. If $P$ is a subset of $[|u|]$, $u_{|P}$
is the subword of $u$ consisting in the letters of $u$ at the positions
specified by the elements of $P$. If $u$ is a word of integers and $k$
is an integer, we denote by $u[k]$ the word obtained by incrementing by
$k$ all letters of $u$. The {\em shuffle} of two words $u$ and $v$ is
the set recursively defined by
$u \shuffle \epsilon = u = \epsilon \shuffle u$ and
$ua \shuffle vb = (u \shuffle vb)a \cup (ua \shuffle v)b$, were $a$ and
$b$ are letters. A word $u$ is a {\em square} if there exists a word $v$
such that $u$ belongs to $v \shuffle v$.

@@

Let us now provide some definitions about graphs and matching that are used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph.
For an edge $e = \{u,v\}$, we shall occasionally say that $u$ and $v$
are \emph{involved} in $e$.
In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$
(\emph{i.e.}, $\mathcal{M}$ avoids \MWithContainment).

@@

This section is devoted to proving hardness of recognizing square
permutations. In the same way as happens with words, we shall use a
linear graph framework where deciding whether a permutation is a square
reduces to computing some specific matching in the associated linear
graph~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}. It reads as follows
(see Fig.\ref{fig:example containment-free matching}).

@@

\begin{figure}[ht!]
  \centering
  \begin{tikzpicture}[scale=0.5,inner sep=2pt,node distance=1cm]
    \draw node (K) {$K_\pi$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=K]   (U01) {$2$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U01] (U02) {$A$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U02] (U03) {$4$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U03] (U04) {$7$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U04] (U05) {$1$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U05] (U06) {$9$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U06] (U07) {$5$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U07] (U08) {$3$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U08] (U09) {$8$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U09] (U10) {$6$};
    \draw [thick]
      (U01.north) .. controls ($ (U01.north) + (0,2.25) $) and ($ (U03.north) + (0,2.25) $) .. (U03.north);
    \draw [thick]
      (U02.north) .. controls ($ (U02.north) + (0,2.25) $) and ($ (U06.north) + (0,2.25) $) .. (U06.north);
    \draw [thick]
      (U04.north) .. controls ($ (U04.north) + (0,2.5) $) and ($ (U07.north) + (0,2.5) $) .. (U07.north);
    \draw [thick]
      (U05.north) .. controls ($ (U05.north) + (0,2.5) $) and ($ (U08.north) + (0,2.5) $) .. (U08.north);
    \draw [thick]
      (U09.north) .. controls ($ (U09.north) + (0,2.5) $) and ($ (U10.north) + (0,2.5) $) .. (U10.north);
  \end{tikzpicture}
  \caption{\label{fig:example containment-free matching}%
    The linear graph $K_\pi$ associated to the permutation
    $\pi = 2A47195386$ together with
    a containment-free perfect matching $\mathcal{M}$
    such that for any two distinct edges $(a, b)$ and
    $(a', b')$ in $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only
    if $\pi(b) < \pi(b')$.
    From $\mathcal{M}$, it follows that $\pi$ is square as it is
    the shuffle of $2A718$ and $49536$ that are both
    order-isomorphic to $25314$.
  }% end caption
\end{figure}

@@

  Let $\pi \in S_n$ and $\sigma \in S_k$ be two arbitrary permutations.
  Define
  \begin{align*}
  N_4 &= 2(2n + k + 2) + 1  = 4n + 2k + 5 \\
  N_3 &= 2(2N_4 + 2n + 2k + 4) + 1 = 20n + 12k + 9 \\
  N_2 &= 2(2N_3 + 2N_4 + 2n + 2k + 4) + 1 \\
  N_1 &= 2(2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4) + 1\text{.}
  \end{align*}
  Notice that $N_1, N_2, N_3$ and $N_4$ are polynomial in $n$.
  The crucial properties are that
  (i) $N_1, N_2, N_3$ and $N_4$ are odd integers
  and
  (ii) $N_i > \left(\sum_{i < j \leq 4} 2N_j\right) + 2n + 2k + k$
  for every $1 \leq i \leq 4$.

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
Given two permutations $\pi$ and $\sigma$, how hard is the problem of
deciding whether $\sigma$ is a square root of $\pi$?
\smallskip

>>>>

%%
%%
%% ---- Introduction ----
%%
\section{Introduction}
\label{section:Introduction}
The {\em shuffle product}, denoted $\shuffle$, is a very famous operation on words
first defined by Eilenberg and Mac Lane~\cite{Eilenberg:MacLane:1953}.
Given three words $u$, $v_1$, and $v_2$, $u$ is said to be a \emph{shuffle}
of $v_1$ and $v_2$ if it can be formed by interleaving the letters from
$v_1$ and $v_2$ in a way that maintains the left-to-right ordering of the
letters from each word.
Besides purely combinatorial questions,
the shuffle product of words leads to the following computational ones:
\begin{enumerate}
    \item \label{item:problem_1}
    Given two words $v_1$ and $v_2$, compute the set $v_1 \shuffle v_2$.
    \item \label{item:problem_2}
    Given three words $u$, $v_1$, and $v_2$, decide if $u$ is a shuffle
    of $v_1$ and $v_2$.
    \item \label{item:problem_3}
    Given words $u$, $v_1$, \dots, $v_k$, decide if $u$ is in
    $v_1 \shuffle \dots \shuffle v_k$.
    \item \label{item:problem_4}
    Given a word $u$, decide if there is a word $v$ such that $u$ is
    in $v \shuffle v$.
\end{enumerate}
Even if these problems seem similar, they are very different in terms
of complexity. Let us now review some facts about these. In what follows,
$n$ denotes the size of $u$ and $m_i$ denotes the size of each $v_i$.
A solution to Problem~\ref{item:problem_1} can be computed in
% \begin{equation}
%     O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)
% \end{equation}
$O\left((m_1 + m_2) \; \binom{m_1 + m_2}{m_1}\right)$
time~\cite{Spehner:TCS:1986}. An improvement and a generalization
of~Problem~\ref{item:problem_1} has been proposed
in~\cite{Allauzen:IGM:2000}, where it is proved that given words
$v_1$, \dots, $v_k$, the iterated shuffle
$v_1 \shuffle \dots \shuffle v_k$ can be computed in
% \begin{equation}
%     O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)
% \end{equation}
$O\left(\binom{m_1 + \dots + m_k}{m_1, \dots, m_k}\right)$
time. Problem~\ref{item:problem_2} is in \Pclass; it is indeed a
classical textbook exercise to design an efficient dynamic programming
algorithm solving it. It can be tested in $O\left(n^2 / \log(n)\right)$
time~\cite{Leeuwen:Nivat:IPL:1982}. To the best of our knowledge, the
first $O(n^2)$ time algorithm for this problem appeared
in~\cite{Mansfield:DAM:1983}. This algorithm can easily be extended to
check in polynomial-time whether or not a word is in the shuffle of any
fixed number of given words. Nevertheless, Problem~\ref{item:problem_3}
is \NPC~\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}. This
remains true even if the ground alphabet has size
$3$~\cite{Warmuth:Haussler:JCSS:1984}. Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that
Problem~\ref{item:problem_3} remains \NPC even if all the words $v_i$,
$i \in [k]$, are identical, thereby proving that, for two words $u$ and
$v$, it is \NPC to decide whether or not $u$ is in the iterated shuffle
of $v$. Again, this remains true even if the ground alphabet has size $3$.
Let us now finally focus on Problem~\ref{item:problem_4}. It is shown
in~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013} that it is \NPC to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that $u$
is a shuffle of $v$ with itself. Hence, Problem~\ref{item:problem_4}
is \NPC.
\smallskip

@@

We follow the usual terminology on words~\cite{ChoffrutKarhumaki1997}.
Let us recall here the most important ones. Let $u$ be a word. The
length of $u$ is denoted by $|u|$. The {\em empty word}, the only word
of null length, is denoted by $\epsilon$. We denote by $\widetilde{u}$
the {\em mirror image} of $u$, that is the word
$u_{|u|} u_{|u| - 1} \dots u_1$. If $P$ is a subset of $[|u|]$, $u_{|P}$
is the subword of $u$ consisting in the letters of $u$ at the positions
specified by the elements of $P$. If $u$ is a word of integers and $k$
is an integer, we denote by $u[k]$ the word obtained by incrementing by
$k$ all letters of $u$. The {\em shuffle} of two words $u$ and $v$ is
the set recursively defined by
$u \shuffle \epsilon = \{u\} = \epsilon \shuffle u$ and
$ua \shuffle vb = (u \shuffle vb)a \cup (ua \shuffle v)b$, were $a$ and
$b$ are letters. A word $u$ is a {\em square} if there exists a word $v$
such that $u$ belongs to $v \shuffle v$.

@@

Let us now provide some definitions about graphs and matching that are used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph.
For an edge $e = \{u,v\}$, we shall occasionally say that $u$ and $v$
are \emph{involved} in $e$.
In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$
(\emph{i.e.}, $\mathcal{M}$ avoids~\MWithContainment).

@@

This section is devoted to proving hardness of recognizing square
permutations. In the same way as happens with words, we shall use a
linear graph framework where deciding whether a permutation is a square
reduces to computing some specific matching in the associated linear
graph~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}. It reads as follows
(see Fig.\ref{fig:example containment-free matching}).
\begin{figure}[ht!]
  \centering
  \begin{tikzpicture}[scale=0.5,inner sep=2pt,node distance=1cm]
    \draw node (K) {$K_\pi$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=K]   (U01) {$2$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U01] (U02) {$A$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U02] (U03) {$4$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U03] (U04) {$7$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U04] (U05) {$1$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U05] (U06) {$9$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U06] (U07) {$5$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U07] (U08) {$3$};
    \draw node [draw, circle, minimum size=.55cm, fill=black!10] [right of=U08] (U09) {$8$};
    \draw node [draw, circle, minimum size=.55cm] [right of=U09] (U10) {$6$};
    \draw [thick]
      (U01.north) .. controls ($ (U01.north) + (0,2.25) $) and ($ (U03.north) + (0,2.25) $) .. (U03.north);
    \draw [thick]
      (U02.north) .. controls ($ (U02.north) + (0,2.25) $) and ($ (U06.north) + (0,2.25) $) .. (U06.north);
    \draw [thick]
      (U04.north) .. controls ($ (U04.north) + (0,2.5) $) and ($ (U07.north) + (0,2.5) $) .. (U07.north);
    \draw [thick]
      (U05.north) .. controls ($ (U05.north) + (0,2.5) $) and ($ (U08.north) + (0,2.5) $) .. (U08.north);
    \draw [thick]
      (U09.north) .. controls ($ (U09.north) + (0,2.5) $) and ($ (U10.north) + (0,2.5) $) .. (U10.north);
  \end{tikzpicture}
  \caption{\label{fig:example containment-free matching}%
    The linear graph $K_\pi$ associated to the permutation
    $\pi = 2A47195386$ together with
    a containment-free perfect matching $\mathcal{M}$
    such that for any two distinct edges $(a, b)$ and
    $(a', b')$ in $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only
    if $\pi(b) < \pi(b')$.
    From $\mathcal{M}$, it follows that $\pi$ is square as it is in
    the shuffle of $2A718$ and $49536$, both being
    order-isomorphic to $25314$.
  }% end caption
\end{figure}

@@

@@

  Let $\pi \in S_n$ and $\sigma \in S_k$ be two arbitrary permutations.
  Define
  \begin{align*}
  N_4 &= 2(2n + 2k + 4) + 1  = 4n + 4k + 9 \\
  N_3 &= 2(2N_4 + 2n + 2k + 4) + 1 = 20n + 20k + 45 \\
  N_2 &= 2(2N_3 + 2N_4 + 2n + 2k + 4) + 1 = 100n + 100k + 225 \\
  N_1 &= 2(2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4) + 1 = 1000n + 1000k  + 1325\text{.}
  \end{align*}
  Notice that $N_1, N_2, N_3$ and $N_4$ are polynomial in $n$.
  The crucial properties are that
  (i) $N_1, N_2, N_3$ and $N_4$ are odd integers
  and
  (ii) $N_i > \left(\sum_{i < j \leq 4} 2N_j\right) + 2n + 2k + k$
  for every $1 \leq i \leq 4$.

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
deciding whether $\sigma$ is a square root of~$\pi$?
